Mergeable Types

### 2017-02-28 paper/evaluation.tex

In this section, we present an evaluation of \name programming model with the
aim of assessing its practical utility for describing geo-distributed
eventually consistent applications. In particular, we are interested in the
horizontal scalability of applications built using \name. Typical distributed
eventually consistent data stores use custom synchronization and dissemination
protocols for transferring updates between replicas. In contrast, \name is
realized over Irmin~\cite{irmin} which uses Git transfer protocol~\cite{..}, a
protocol not designed with distributed data stores in mind. Hence, we also
evaluate the performance of synchronization and merging.

>>>>

In this section, we present an evaluation of \name programming model with the
aim of assessing its practical utility for describing geo-distributed
eventually consistent applications. In particular, we are interested in the
horizontal scalability of applications built using \name. Typical distributed
eventually consistent data stores use custom synchronization and dissemination
protocols for transferring updates between replicas. In contrast, \name is
realized over Irmin~\cite{irmin} which uses Git transfer
protocol~\cite{git-tp}, a protocol not designed with distributed data stores in
mind. Hence, we also evaluate the performance of synchronization and merging.

### 2017-02-28 paper/semantics.tex
edits

\subsection{Operational Semantics}

We formalize our ideas in the context of a lambda calculus ($\lang$)
shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
constants, and \name primitives composed using the lambda combinator.
For brevity, we use short names for \name primitives: \C{run} for
\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
simplify the technical development, \name's \C{sync\_next\_version}
operation is broken down into two primitives - \C{push} and \C{pull},
which can be composed to get the desired effect of the original:
\begin{smathpar}
  \C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)
\end{smathpar}
The semantics of\C{get\_current\_version} is subsumed by \C{pull},
hence elided.  Values ($v$) are constants and lambda abstractions.  A
program ($p$) is a parallel composition of threads, where each thread
is an expression ($s$) indexed by the corresponding thread identifier
($t$).

@@

$E$ is defined to evaluate the first argument of a \C{run} expression
to a value that constitutes an initial version (recall that \C{run}
models \name's \C{with\_init\_version\_do}). A program evaluation
context non-deterministically picks an extant thread to evaluate. The
administrative rule that relates transitions of holes to transitions
of expressions and programs is straightforward, and hence elided. The
remaining core reduction rules are presented in
Fig.~\ref{fig:opsem}. For brevity, we write $H(t\mapsto (v,f))$ to
denote the proposition that $H$ maps $t$ to $(v,f)$. The notation $H[t
  \mapsto (v,f)]$ denotes the extension of $H$ with the binding $t
\mapsto (v,f)$, as usual.

@@

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The versions are merged only if they are \emph{safely
mergeable} (denoted $\under{H}{v' \mbleto v}$) - a notion that we will
explain shorty. The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{lowest common ancestor} (LCA) for the
latest versions on the given pair of branches. The formal definition
of LCA is given below. The \rulelabel{E-Pull-Wait} and
\rulelabel{E-Pull} rules thus let a thread synchronize with other
distributed threads manipulating versions of replicas in multiple
steps before returning the result of the \C{pull}. Since \C{sync} is a
composition of \C{push} and \C{pull}, its behavior can be explained
thus: \C{sync} pushes the given value onto the current (local) branch,
merges a (possibly empty) subset of concurrent branches into the local
branch, and returns the result.  (The operation is not guaranteed to
synchronize with all concurrent branches because not all such branches
may be available.)

@@

\begin{definition} [\bfseries Ancestor]
Version $v_1$ is a ancestor of version $v_2$ under a history
$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
following is true:
\begin{itemize}
  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
  $v_1$,
  \item There exists a branch $b$ in $H$ that contains $(v_2,
  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item There exists a branch $b$ in $H$ that contains
  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item $v_1 = v_2$, or $v_1$ is transitively a ancestor of
  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
  \under{H}{v \preceq v}$
\end{itemize}
\end{definition}

@@

The function \C{lca} used in the premise of \rulelabel{E-Pull-Wait}
computes the LCA of two branches, and assumes it to be unique.  But,
observe that the definition of LCA doesn't guarantee its uniqueness;
it has to be enforced explicitly.  The semantics enforces the
uniqueness of LCA by constraining the branching structure via the
\rulelabel{E-Pull-Wait} rule. The rule merges version $v'$ into
version $v$ only if they are safely mergeable ($\under{H}{v' \mbleto
v}$), i.e., they have a single LCA, and merging both versions does not
lead to a case where a branch (i.e., its latest version) has multiple
LCAs with rest of the branches, hence gets \emph{stuck}. Such nuances
of LCA, and the guarantees provided by the system notwithstanding
these nuances, are discussed in the Sec.~\ref{sec:meta}.

>>>>

\subsection{Operational Semantics}
\label{sec:opsem}

We formalize our ideas in the context of a lambda calculus ($\lang$)
shown in Fig.~\ref{fig:opsem}. Expressions of $\lang$ are variables,
constants, and \name primitives composed using the lambda combinator.
For brevity, we use short names for \name primitives: \C{run} for
\C{with\_init\_version\_do}, and \C{fork} for \C{fork\_version}. To
simplify the technical development, \name's \C{sync\_next\_version}
operation is broken down into two primitives - \C{push} and \C{pull},
which can be composed to get the desired effect of the original:
$\C{sync}\;x \;=\; (\lambda y.\pull)\; (\push\,x)$.  The semantics
of\C{get\_current\_version} is subsumed by \C{pull}, hence elided.
Values ($v$) are constants and lambda abstractions.  A program ($p$)
is a parallel composition of threads, where each thread is an
expression ($s$) indexed by the corresponding thread identifier ($t$).

@@

$E$ is defined to evaluate the first argument of a \C{run} expression
to a value that constitutes an initial version (recall that \C{run}
models \nameMonad's \C{with\_init\_version\_do}). A program evaluation
context non-deterministically picks an extant thread to evaluate. The
administrative rule that relates transitions of holes to transitions
of expressions and programs is straightforward, and hence elided. The
remaining core reduction rules are presented in
Fig.~\ref{fig:opsem}. For brevity, we write $H(t\mapsto (v,f))$ to
denote the proposition that $H$ maps $t$ to $(v,f)$. The notation $H[t
  \mapsto (v,f)]$ denotes the extension of $H$ with the binding $t
\mapsto (v,f)$, as usual.

@@

The semantics non-deterministically chooses \rulelabel{E-Pull} or
\rulelabel{E-Pull-Wait} rules to reduce a \C{pull} expression. The
\rulelabel{E-Pull} rule reduces \C{pull} to \C{()}, and returns the
latest version on the current branch. The \rulelabel{E-Pull-Wait} rule
can be thought of as a stutter step; it doesn't reduce \C{pull}, but
updates the branching structure by merging (the latest version of) a
concurrent branch ($H(t')$) into (the latest version of) the current
branch ($H(t)$), and extending the current branch with the merged
version ($v_m$). The versions are merged only if they are \emph{safely
mergeable} (denoted $\under{H}{v' \mbleto v}$) - a notion that we will
explain shorty. The new version is tagged with a \C{MERGE} tag that,
like a \C{FORK} tag, records its origin. The rule assumes a function
\C{lca} that computes the \emph{lowest common ancestor} (LCA) for the
latest versions on the given pair of branches. The formal definition
of LCA is given below. The \rulelabel{E-Pull-Wait} and
\rulelabel{E-Pull} rules thus let a thread synchronize with other
distributed threads manipulating versions of replicas in multiple
steps before returning the result of the \C{pull}. Since \C{sync} is a
composition of \C{push} and \C{pull}, its behavior can be explained
thus: \C{sync} pushes the given value onto the current (local) branch,
merges a (possibly empty) subset of concurrent branches into the local
branch, and returns the result.  (The operation is not guaranteed to
synchronize with all concurrent branches because not all such branches
may be available, because of e.g., network partitions.)

@@

\begin{definition} [\bfseries Ancestor]
Version $v_1$ is a ancestor of version $v_2$ under a history
$H$ (written $\under{H}{v_1 \preceq v_2}$) if and only if one of the
following is true:
\begin{itemize}
  \item There exists a branch $b$ in $H$ (i.e., $\exists(t\in
  dom(H)).\,H(t) = b$) in which $v_2$ immediately succeeds
  $v_1$,
  \item There exists a branch $b$ in $H$ that contains $(v_2,
  \C{FORK}\; (v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item There exists a branch $b$ in $H$ that contains
  $(v_2, \C{MERGE}\;(v_1,f_1)::b_1)$, for some $f_1$ and $b_1$,
  \item $v_1 = v_2$, or $v_1$ is transitively an ancestor of
  $v_2$, i.e., $\exists v.~ \under{H}{v_1 \preceq v} \conj
  \under{H}{v \preceq v}$
\end{itemize}
\end{definition}

@@

The function \C{lca} used in the premise of \rulelabel{E-Pull-Wait}
computes the LCA of two branches, and assumes it to be unique.  But,
observe that the definition of LCA doesn't guarantee its uniqueness;
it has to be enforced explicitly.  The semantics enforces the
uniqueness of LCA by constraining the branching structure via the
\rulelabel{E-Pull-Wait} rule. The rule merges version $v'$ into
version $v$ only if they are safely mergeable ($\under{H}{v' \mbleto
  v}$), i.e., they have a single LCA, and merging both versions does
not lead to a case where a branch (i.e., its latest version) has
multiple LCAs with remaining branches, andd hence gets
\emph{stuck}. Such nuances of LCA, and the guarantees provided by the
system notwithstanding these nuances, are discussed in 
Sec.~\ref{sec:meta}.
